---
title: "Notation, Functions, and Limits"
subtitle: "MathCamp 2026 — Session 1"
---

## Introduction & Logistics {.smaller}

-   Introductions
-   Math camp goals, logistics, and schedule
-   What are quantitative methods? Quantitative vs. qualitative methods

$\beta = \alpha$

## Variables & Measures {.smaller}

-   Variables and constants
-   Sets and sample spaces
-   Unions, intersections, subsets (set notation)
-   Levels of measurement, and why it matters

## Rules of Algebra {.smaller}

-   Basic properties of arithmetic
-   Order of operations
-   Inequalities
-   Exponents, logarithms, and roots

## Functions {.smaller}

-   Functions and their characteristics
-   Function composition
-   Graph examples: linear, quadratic, exponential functions
-   Functions and theorizing; Anscombe's quartet

## Solving Equations {.smaller}

-   Special products and factoring
-   Quadratics and special products
-   Quadratic formula, checking work graphically

## Series, Sequences, and Limits

-   Sequences and series
-   Limits (including piecewise function examples)

## Dataset preparation {.smaller}

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
load("../chapters/02-probability/data/JPRdata_Haim.rda")
COW_country_codes <- read_csv("../chapters/02-probability/data/COW-country-codes.csv")

JPRdata_joined <- left_join(JPRdata, COW_country_codes, by = c("ccode_e" = "CCode"))
JPRdata_joined <- JPRdata_joined %>%
  rename(StateAbb_export = StateAbb,
         StateNme_export = StateNme) %>%
  relocate(
    StateAbb_export,
    StateNme_export,
    .after = ccode_e
  ) 

JPRdata_joined <- left_join(JPRdata_joined, COW_country_codes, by = c("ccode_a" = "CCode"))
JPRdata_joined <- JPRdata_joined %>%
  rename(StateAbb_import = StateAbb,
         StateNme_import = StateNme) %>%
  relocate(
    StateAbb_import,
    StateNme_import,
    .after = ccode_a
  ) 

JPRdata_cut <- JPRdata_joined %>%
  filter(year > 2000)

JPRdata_select <- JPRdata_cut%>%
  select(year,
         ccode_e,
         StateAbb_export,
         StateNme_export,
         ccode_a,
         StateAbb_import,
         StateNme_import,
         logflow,
         shareally,
         MID,
         loggdp.e,
         loggdp.a,
         comcur,
         sscore,
         region,
         bilatno,
         multino)

```

## Dataset {.smaller}

This sample dataset is adapted from the replication data accompanying:

> Haim, Dotan. 2016. *Alliance Networks and Trade: The Effect of Indirect Political Alliances on Bilateral Trade Flows*. Journal of Peace Research.

The original replication files and codebook are available through the Journal of Peace Research replication archive and the author's website. This version retains a subset of variables for instructional purposes and adds country names and abbreviations by merging the Correlates of War (COW) country code list.

## Variables {.smaller}

| Variable          | Description                                                                                                                                                                                                                                        |
|:------------------------------|:---------------------------------------|
| `year`            | Year of the observation.                                                                                                                                                                                                                           |
| `ccode_e`         | Correlates of War (COW) numeric country code for the exporting (ego) state.                                                                                                                                                                        |
| `StateAbb_export` | Three-letter country abbreviation for the exporting state (added from the COW country code list).                                                                                                                                                  |
| `StateNme_export` | Country name for the exporting state (added from the COW country code list).                                                                                                                                                                       |
| `ccode_a`         | Correlates of War (COW) numeric country code for the importing (alter) state.                                                                                                                                                                      |
| `StateAbb_import` | Three-letter country abbreviation for the importing state (added from the COW country code list).                                                                                                                                                  |
| `StateNme_import` | Country name for the importing state (added from the COW country code list).                                                                                                                                                                       |
| `logflow`         | Natural logarithm of exports from the exporting state to the importing state, measured in millions of standardized U.S. dollars. Data originate from the CEPII Trade dataset.                                                                      |
| `shareally`       | Number of alliance ties jointly shared by both members of a dyad, adjusted for allies unique to the importing state. This variable measures indirect alliance embeddedness.                                                                        |
| `MID`             | Indicator for whether the two states were involved in a Militarized Interstate Dispute (MID) during the observation year.                                                                                                                          |
| `loggdp.e`        | Natural logarithm of the exporting state's gross domestic product (GDP), measured in millions of standardized U.S. dollars.                                                                                                                        |
| `loggdp.a`        | Natural logarithm of the importing state's gross domestic product (GDP), measured in millions of standardized U.S. dollars.                                                                                                                        |
| `comcur`          | Indicator for whether the two states share a common currency.                                                                                                                                                                                      |
| `sscore`          | Affinity (S-) score measuring similarity in the two states' United Nations General Assembly voting behavior.                                                                                                                                       |
| `region`          | Geographic region of the exporting state. Categories include East Asia and Pacific, South Asia, Middle East and North Africa, Sub-Saharan Africa, Europe and Central Asia, Latin America and Caribbean, North/Central America, and Western Europe. |
| `bilatno`         | Indicator for whether the two states are partners in at least one bilateral alliance.                                                                                                                                                              |
| `multino`         | Indicator for whether the two states are partners in at least one multilateral alliance.                                                                                                                                                           |

## Notes {.smaller}

-   Each observation represents a **directed country dyad-year**, where an exporting state trades with an importing state.
-   Variables beginning with `.e` refer to characteristics of the exporting (ego) state, whereas variables beginning with `.a` refer to the importing (alter) state.
-   `StateAbb_*` and `StateNme_*` were added to improve readability and are not part of the original Haim (2016) replication dataset.

```{r}
head(JPRdata)
```

##  {.center}

**Programming lecture:** [webpage chapter](../chapters/01-notation-functions-limits/programming.html)
